---
title: "Functionals"
author: "Mark Ruddy"
date: "5 March 2017"
output: html_document
---

```{r load-packages, include=FALSE}
library(tidyverse)
```

## Functionals introduction

Working through HW's Advanced-R tutorial on functionals: http://adv-r.had.co.nz/Functionals.html 

A functional:

> A function that takes a function as an input and returns a vector as output.

eg.,

```{r functional-example}
randomise <- function(f) f(runif(1e3))
glimpse(randomise(sum))
```

`lapply()`, `apply()` and `tapply()` are the most frequently use functionals.

Functionals are often used as alternatives to 'for loops'. 

1. For loops have a reputation for beign slow (although not always).
2. Using a functional is more expressive than relying on a for loop.
  * Each functional tailored to a specific task: know immediately why it's being used.
3. Eliminates having to write boilerplate for for-loop and so concentrate on function being applies.
4. Functionals reduce bugs (well tested) and are efficient.
5. Many functionals have been written in C with performance in mind (although not always).


## lapply()

See http://adv-r.had.co.nz/Functionals.html#lapply

`lapply()` 

1. Takes a function and a list.
2. Creates container for output.
3. Applies function to each element in the list.
4. Returns results as a list.

It's written in C, is the simplest functional, and is a building block for many other functionals.


### Looping patterns

There are three ways to loop over a vector.

1. Over elements: `for (x in xs)`
2. Over numeric indices: `for (x in seq_along(xs))`
3. Over names: `for (nm in names(xs))`

Usually 1 is a bad choice: saving output usually is by exending a data structure - eg extending a vector `res <- c(res, fun(x))`), which is inefficient. Extending a vector requires R to copy all existing elements, the add to it. Better to create the space needed for output first, then fill it.

eg

```{r index-for-loop-example}
res <- numeric(length(xs))
for (i in seq_along(xs)) {
  res[i] <- sqrt(xs[i])
}
```

### Using lapply()

Three basic ways:

1. With elements: `lapply(xs, function(x) {})`
2. With indices: `lapply(seq_along(xs), function(i) {})`
3. With names: `lapply(names(xs), function(nm) {})`

In most cases use case 1 - `lapply()` sorts out the output. 

If the position or name of the element is needed, use cases 2 or 3.

## Exercises

1. Why are the following two invocations of `lapply()` equivalent?

```{r Ex1-lapply}
trims <- c(0, 0.1, 0.2, 0.5)
x <- rcauchy(100)

#a
lapply(trims, function(trim) mean(x, trim = trim))

#b
lapply(trims, mean, x = x)
```

The second parameter of `lapply` can take a name of a function found via `match.fun`. Thus `lapply(trims, 'mean', x = x)` also works, even though Hadley doesn't mention that.

`lapply` will then automatically apply the function from the second parameter to the first parameter.  Though, by default it will assume the first argument to `lapply` will be the first argument to the function unless you specify otherwise in the third argument of `lapply`.

Here, mean takes three args: x, trim, and na.rm.  `lapply(trims, mean)` would do `mean(trims)` or `mean(x = trims)`.  But by specifying that `x = x`, lapply then knows that the x value for mean will be the x variable, and then passes trims as the second argument to `mean`.

(With reference to https://github.com/peterhurford/adv-r-book-solutions/blob/master/09_functionals/01_lapply/exercise1.r)


2. The function below scales a vector so it falls in the range [0, 1]. How would you apply it to every column of a data frame? How would you apply it to every numeric column in a data frame?

```{r Ex2-df-q}
scale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```


```{r Ex2-df-ans}
str(iris)
names(iris)

lapply(mtcars, scale01)
# or with selection for numeric columns
lapply(iris[sapply(iris, is.numeric)], scale01)
```


3. Use both for loops and lapply() to fit linear models to the mtcars using the formulas stored in this list:

```{r Ex3-loop-q}
formulas <- list(
  mpg ~ disp,
  mpg ~ I(1 / disp),
  mpg ~ disp + wt,
  mpg ~ I(1 / disp) + wt
)
```

```{r Ex3-loop-a} 

# For loop
fits1 <- vector("list", length = length(formulas))
i <- 1
for (formula in formulas) { fits1[[i]] <- lm(formula, data = mtcars); i <- i + 1 }
fits1

# lapply
fits2 <- lapply(formulas, lm, data = mtcars)

```



4. Fit the model mpg ~ disp to each of the bootstrap replicates of mtcars in the list below by using a for loop and lapply(). Can you do it without an anonymous function?

```{r Ex4-fit-q}
bootstraps <- lapply(1:10, function(x) {
  rows <- sample(1:nrow(mtcars), rep = TRUE)
  mtcars[rows, ]
})
```

```{r Ex4-fit-a}
# For loop
mod1 <- vector("list", length = length(bootstraps))
i <- 1
for(boot in bootstraps) { 
  result <- lm(mpg ~ disp, data = boot)
  mod1[[i]] <- result
  i <- i+1
  }
mod1

# lapply
mod2 <- lapply(bootstraps, lm, formula = mpg ~ disp)
```


5. For each model in the previous two exercises, extract R2 using the function below.

```{r Ex5-2funs-q}
rsq <- function(mod) summary(mod)$r.squared
```


```{r Ex5-2funs-q}
lapply(mod2, rsq)

# extract R2 for all
all.fits <- list(fits1,fits2,mod1,mod2)
all.r2 <- lapply(all.fits, function(fit) lapply(fit, rsq))
```

Clean up
```{r}
rm(list = ls())
```


## For loop functionals: friends of lapply()

Common looping functions are *already implemented* in existing base functionals.

### Vector output: `sapply` and `vapply`

Both produce atomic vectors as outputs but `sapply` assumes output and so is more prone to errors of input is not as expected. `vapply` takes more arguments to control implementation and output.

`sapply` is just `lapply` with the list converted to a vector.

`vapply` is `lapply` with appropriate output (eg matrix) instead of a list.

Use `sapply` when interactive exploring/investigation - it silently returns a list if the FUN outputs results of different lengths. Use `vapply` when including within functions as it throws an error if FUN results are different lengths.

## Multiple outputs: `Map` (and `mapply`)

These allow iteration over two lists at the same time, eg is calculating weighted means.

`Map` takes a function as the first argument, cf `lapply` et al. 

`Map` is the equivalent of `mapply` but with `simplify=FALSE`

```{r test1}
xs <- replicate(5, runif(10), simplify = FALSE)
ws <- replicate(5, rpois(10, 5) + 1, simplify = FALSE)

unlist(Map(weighted.mean, xs, ws))
```

## Bespoke functionals

Try and spot the for loop structures wthin any novel problems encountered, relate them to existing functionals and make use these functionals in any new functions.


## Parallelisation


`parallel::mclapply` and `parallel::mcMap` are versions of `lapply` and `Map` that can use multiple cores. This is because each iteration with these functionals is independent of others, ie not order dependent.

### Exercises

1. Use vapply() to:
  a. Compute the standard deviation of every column in a numeric data frame.
  b. Compute the standard deviation of every numeric column in a mixed data frame. (Hint: you’ll need to use vapply() twice.)

```{r ex1-vapply}
# with vapply we have to be explicit obout the output. Focus on FUN.VALUE argument, because all values of FUN must have same length and type.

# 1a
vapply(mtcars, sd, numeric(1))

# 1b
vapply(iris[vapply(iris, is.numeric, logical(1))], sd, numeric(1))
```


2. Why is using sapply() to get the class() of each element in a data frame dangerous?

If an input has one class `sapply` will output a vector, but with inputs of multiple classes the output will be a list. This could be unexpected.
Also, `sapply` returns an empty list when supplied with an empty list - not a logical vector. 


3. The following code simulates the performance of a t-test for non-normal data. Use sapply() and an anonymous function to extract the p-value from every trial.

```{r Ex3-sapply-test-q}
trials <- replicate(
  100, 
  t.test(rpois(10, 10), rpois(7, 10)),
  simplify = FALSE
)
trials[[1]]$p.value


sapply(trials, function(x) x$p.value)
```


4. What does replicate() do? What sort of for loop does it eliminate? Why do its arguments differ from lapply() and friends?

`replicate` repeatedly evaluates an expression. It eliminates for looping over an expression. It differs from `lapply` and other related functionals in that it takes no data and it repeated calls an expression `n` times.










































